---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Introduction {#intro}

Statistics  
  - allows us to reason from data
  - need a mathematical framework
  
## Things to come:
  - simple linear regression
  - identify a line that 'best' fits the data

```{r load-packages, warning = FALSE}
library(stfspack)
library(tidyverse)
theme_set(theme_bw(base_size = 12) + 
            theme(strip.background = element_blank(), 
                  panel.grid = element_blank())) 
```

```{r m1, fig.cap = 'Fertilizer consumption and cereal yield in 11 sub-Saharan African countries', out.width='80%', fig.asp=.75, fig.align='center'}
mod.fit <- lm(y1 ~ x1, data = anscombe)
anscombe %>% 
  ggplot(aes(x1, y1)) + 
  geom_point() 
```

```{r m2, fig.cap = 'The agriculture data with a line of best fit from the simple linear regression', out.width='80%', fig.asp=.75, fig.align='center'}
mod.fit <- lm(y1 ~ x1, data = anscombe)
anscombe %>% 
  ggplot(aes(x1, y1)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```
  
```{r view-mod}
summary(mod.fit)
```

```{r m3, fig.cap = 'The data underlying the analysis of the variables y3 and x3 in the anscombe data set'}
mod.fit2 <- lm(y3 ~ x3, data = anscombe)
anscombe %>% 
  ggplot(aes(x3, y3)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

```{r view-mod2}
summary(mod.fit2)
```

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```
