---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```

```{r load-packages, include = FALSE}
library(tidyverse)
#library(stfspack)
theme_set(theme_bw(base_size = 12) + 
            theme(strip.background = element_blank(), 
                  panel.grid = element_blank())) 
```

# Bayesian estimation and inference {#bayesian}

Parameters

  - Frequentist view: parameters are fixed
  - Bayesian view: parameter are random

> Posterior distribution: the probability distribution of the parameter(s) being estimated, given the data observed

> Prior distribution: the probability distribution function used to 'describe' the parameter(s) before any data are observed

If we have a prior, we can calculate the posterior $f(\theta | D = d)$ as follows:

$$
\begin{aligned}
f(\theta | D = d) =& ~ \frac{f_D(d|\theta) ~ f_{\theta}(\theta)} {f_D(d)} \\
\end{aligned}
$$
where the likelihood $f_D(d|\theta)$ is multiplied by the prior $f_{\theta}(\theta)$ in the numerator, which is divided by the probability of the data $f_D(d)$. The probability of the data is the probability of observing the data given the parameter, averaged across all parameter values weighted by the probability (or density) of each parameter value:

$$
\begin{aligned}
f_D(d) =& ~ \int_{-\infty}^{\infty} f_D(d|\theta) f_{\theta}(\theta) d\theta\\
\end{aligned}
$$


## How to choose a prior distribution?

## The unscaled posterior, conjugacy, and sampling from the posterior

### Exercise set 10-1

1. 

```{r}
n <- 20
true.mean <- 2
known.sd <- 1
prior.mean <- 0
prior.sd <- 1
set.seed(8675309)
z <- rnorm(n,true.mean,known.sd)
mean(z)
hist(z)
```

a. 

```{r}
post.conj.norm.norm <- function(z, known.sd, prior.mean,
prior.sd){
  xbar <- mean(z)
  post.expec <- (prior.mean / prior.sd^2 + xbar*length(z) /
        known.sd^2)/(1 /   prior.sd^2 + length(z) / known.sd^2)
  post.var <- 1 / (1 /   prior.sd^2 + length(z) / known.sd^2)
  list("posterior.expectation" = post.expec,
"posterior.variance" = post.var)
}

post.conj.norm.norm(z, known.sd, prior.mean, prior.sd)

```

c. 

```{r}
library(MCMCpack)
mn.mod <- MCnormalnormal(z, sigma2 = 1, mu0 = prior.mean, tau20 = prior.sd^2, mc = 10000)
summary(mn.mod)
```

## Bayesian point estimation using Bayes estimators

### Exercise set 10-2

## Bayesian interval estimation using credible intervals

### Exercise set 10-3

## Bayesian 'hypothesis testing' using Bayes factors

### Exercise set 10-4

## Conclusion: Bayesian vs. frequentist methods

## Chapter summary